---
title: "Severe Weather Events Impact on Public Health and Economics Analysis Report"
author: "Tony G. Bolaño"
date: "Sunday, May 24, 2015"
output:
    html_document:
        toc: true
---




## Summary ##

This report analyzes the impact of severe weather events on public health and economics for communities and municipalities. The analysis is based upon the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database which tracks characteristics of major storms and weather events, estimates fatalities, injuries and property damage in the U.S.A. between year 1950 and 2011. More recent years should be considered more complete.




## Basic Settings

By default all code for the analysis is shown in the document and we prevent scientific notation for numbers.

It also load all necessary libraries at top of the analysis.

```{r "Basic settings"}
echo = TRUE
options(scipen = 1)
```




## Data Proccessing ##

### Loading Data from File ###

```{r "Loading data from file"}
# First we need download data
dataurl <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
datafile <- "StormData.csv.bz2"

if (!file.exists(datafile)) {
    download.file(url = dataurl, destfile = datafile)
}

# R can read csv files form bzip compressed files directly so
# we don't need uncompress the file

storm <- read.csv(datafile)
```

### Tidying data ###

With the objective of asking this two questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

Only eight columns of the 37 in the original dataset are relevant to our analysis, in accordance with the definition in [4] LAYOUT Storm Events Database May 2008. 

Column name | Description
---------------------------------------------------------------
 BGN_DATE   | Date the storm event began
 EVTYPE     | Type of storm event. Take note that similar storm events can be listed using different wording e.g. “coastal flood” and “coastal flooding.” Take note of this if you want to run a query grouping by event type.
 FATALITIES | Number directly killed
 INJURIES   | Number directly injured
 PROPDMG    | Property damage in whole numbers and hundredths
 PROPDMGEXP | A multiplier where Hundred (H), Thousand (K), Million (M), Billion (B)
 CROPDMG    | Crop damage in whole numbers and hundredths
 CROPDMGEXP | A multiplier where Hundred (H), Thousand (K), Million (M), Billion (B)


There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.




## Results ##

There should be a section titled Results in which your results are presented.




## References ##


- [1] [Original data file](http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) from Coursera Course Assignment page.
- [2] National Weather Service [Storm Data Documentation](http://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).
- [3] National Climatic Data Center [Storm Events FAQ](http://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf).
- [4] [LAYOUT Storm Events Database May 2008](http://ire.org/media/uploads/files/datalibrary/samplefiles/Storm%20Events/layout08.doc)


